Towards a Planckian standard kilogram

### 2018-07-18 ms.tex
small edits

\abstract
% context heading (optional)
% {} leave it empty if necessary  
{As the kilogram assumes its new definition in terms of $h,c,$ and the
  standard second, it is interesting to think about a possible future
  definition entirely in terms of fundamental constants.
  A kilogram could, in principle, be defined as an integer multiple of
  the Planck mass (or 4594????$\,\times\sqrt{\hbar c/G}$) without
  significantly altering its current value.  The practical hurdle is
  measuring $G$ four orders of magnitude better than currently known,
  in order to fix the four unknown digits.}
% aims heading (mandatory)
{We assess the prospects for measuring $G$ to the desired accuracy
  using an artificial two-body orbit.}
% methods heading (mandatory)
{We consider the simplest configuration, which would be a system
  placed on a Trojan orbit to minimize external gravitational
  perturbations, and enclosed in a cavity chilled to liquid-helium
  temperatures, to suppress radiation pressure.}
% results heading (mandatory)
{Data analysis would involve marginalizing over at least 15~nuisance
  parameters (6~for orbital parameters and 9~for fictitious forces)
  but that, we show, is feasible.}
% conclusions heading (optional), leave it empty if necessary 
{We have not considered costs in this preliminary study, but more
  complicated but more affordable scenarios in Earth orbit merit
  investigation.}

@@

The reform of SI units to use quantum standards is an ongoing ``news
story'' in physics.\footnote{International Bureau of Weights and
  Measures, ``DRAFT 9th edition of the SI Brochure", BIPM.}  At the
heart of the reform are two quantum phenomena that appear on
macroscopic scales: the Josephson effect makes $h/e$ measurable as a
voltage-to-frequency ratio, while the quantum Hall effect makes
$h/e^2$ a standard resistance (the von~Klitzing constant).  These
enable electrical units to be defined in terms of $h$ and $e$.  A
further ingenious instrument, known as a watt balance, precisely
compares electrical and mechanical power, thus enabled mass to be
standardized in terms of $h$.  For a simple introduction to the new
SI, we refer the reader to the article on the ``LEGO watt
balance'' \citep{Chao15}. The proposed redefinition of the kilogram is
an explicit value for $h$ and $c$, but we can think of it as
\begin{equation}
\mathrm{kg} \equiv 1.3563925\ldots \times 10^{50} \, \mathrm{s}^{-1}
                   \times \frac{h}{c^2}
\end{equation}
with the exact value to be decided.  As we can see, the kilogram is
coupled to the definition of a second.  That is not a problem in
practice, as the second is standardized orders of magnitude better
than any other unit.  Nonetheless, it is interesting to consider a
logical further step of decoupling the kilogram from other
definitions.  This could be done by introducing the gravitational
constant, since $\sqrt{hc/G}$ has dimensions of mass.  Historically,
there have been several proposed unit systems that included $G$ in
some way \citep[for a review, see][]{Tomilin1999}, but the idea of
including both gravity and quantum mechanics in a unit system comes
from \cite{Planck99}. In the process of working towards the
radiation law, Planck realized that $c,G,h$ and the Boltzmann constant
$k_B$ (in his notation $G\rightarrow f$, $h\rightarrow b$,
$h/k_B\rightarrow a$) naturally define units for mass, length, time,
and temperature.  Indeed, Planck remarked that such units would be
meaningful even in non-human and extraterrestrial cultures.

A hypothetical Planckian SI unit system would have
\begin{equation} \label{eq:defs}
\begin{aligned}
\mathrm{kg} &= 4.594\ldots \times 10^7
               \left(\frac{\hbar c}{G}\right)^{\!1/2} \\
\mathrm{m}  &= 6.187\ldots \times 10^{34}
               \left(\frac{\hbar G}{c^3}\right)^{\!1/2} \\
\mathrm{s}  &= 1.854\ldots \times 10^{43}
               \left(\frac{\hbar G}{c^5}\right)^{\!1/2} \\
\mathrm{K}  &= 7.058\ldots \times 10^{-33}
               \left(\frac{\hbar c^5}{Gk_B^2}\right)^{\!1/2}
\end{aligned}
\end{equation}
with the numerical coefficients chosen so as to match the present
standards.  The square-rooted expressions are conventionally known as
the Planck mass, Planck length, and so on.  (They are actually
$\sqrt{2\pi}$ smaller than Planck's definitions.)  Extension to
electrical units is straightforward, but we will not discuss it here.
The Planck length, time, and temperature in themselves are extreme
scales, not probed by any known physical process, and presumably the
regime of quantum gravity.  But extreme scales do not harm if they are
only part of some expression, as they are here.  The Planck mass is
small but microscopically measurable: $\sim20\mu\mathrm{g}$.  This is
of the same order as the uncertainty in the prototype kilogram
itself.\footnote{\url{<https://www.bipm.org/utils/common/pdf/cms_m.pdf>}}
Hence the kilogram could be redefined as an integer multiple of the
Planck mass, without significantly perturbing its value.

@@

\subsection{An artificial two-body gravitating system}
Newton's law of gravitation does not only explain why an apple falls
down from a tree, it also explains orbits. If one has two bodies of
masses $m_1$ and $m_2$, on which the only force acting is their
respective gravitational attraction, they will orbit around the common
center of mass, with the relative vector tracing an ellipse
described by Kepler's laws. The third law, relates the
orbital period $P$ to semimajor axis $a$ of the ellipse, as
\begin{equation}\label{eq:kepler3}
P^2 = \frac{4 \, \pi^2}{G(m_1+m_2)}a^3
\end{equation}
with the gravitational constant $G$ appearing in the proportionality
factor.  If one can measure the position of the two masses very
accurately and thus the orbit is known, it is possible to deduce the
gravitational constant.  So what we need is a freely-falling two-body
system. The only way to achieve this over some time, is to go into
space. Thus, one has to design a spacecraft with a cavity that
contains the two masses and a position measurement apparatus. The
cavity should shield the two-body system from environmental
influences. The spacecraft is then put in space, at a location where
tidal forces are minimal.  The stable Lagrange points of the Earth-Sun
system (that is, $L_4$ or $L_5$, also known as the Earth-Trojan
locations) would be suitable.

@@

\textbf{Radiation pressure:} The envelope of the cavity needs to protect the two body-system from the constant bombardment of radiation and charged particles from the Sun. Since the activity of the Sun varies much, the spacecraft must be able to correct its position around the freely falling masses inside the cavity.\\
Furthermore, there is thermal radiation inside the cavity, that could disturb the orbit.
Using the Stefan-Boltzmann law, we can relate radiation pressure to the temperature inside the cavity. If we then specify to balls of diameter $d$, we can express the force on the balls as 
\begin{equation}
F = \frac{2 \pi^6 d^2 {k_\mathrm{B}}^4 T^4}{45 c^3 h^3}.
\end{equation}
As we can see, this force is highly temperature dependent. In order to minimize this influence, one should deploy a temperature as small as possible in the cavity. In fact, if a cooling system based on liquid helium is installed, temperature can be brought down to about 4 K. Such a low temperature results in a force on the balls in the order of $10^{-16}$~N. Compared to the gravitational force, this is orders of magnitude less and thus we don't have to worry about effects of radiation pressure.

@@

\textbf{Centrifugal and tidal forces:}
\label{sec:tidal}
The free fall condition is not possible to achieve for both masses simultaneously since we are in a non-inertial, rotating frame and therefore have to account for centrifugal forces. The centrifugal acceleration $\ddot{\vec{r}}_{\mathrm{Z}}$ is proportional to the distance vector $\vec{r}$ and scales with $\omega^2$, where $\omega$ is the angular velocity of the Sun-Earth reference frame:
\begin{equation}
\label{eq:centrifugalacc}
\ddot{\vec{r}}_{\mathrm{Z}} = \vec{\omega} \times \left( \vec{\omega} \times \vec{r} \right) 
\end{equation}
In addition, there are tidal forces. If one mass is closer to the Sun, the gravitational force on it is slightly bigger than on the other mass. This results in a force perturbation, which is also proportional to $r$. One can approximate the tidal acceleration as
\begin{equation}
\label{eq:tidalacc}
\ddot{\vec{r}}_{\mathrm{\tau}} \approx \frac{2 \, G \, M_\odot}{R^3} \, \vec{r}.
\end{equation}
In this equation $M_\odot$ is the solar mass and $R$ is the radius of the orbit around the Sun. With Kepler's third law one can rewrite the tidal acceleration as a function of the orbital period squared of the Sun-Earth reference frame. Thus it scales the same way as the centrifugal force. To estimate the magnitude of the centrifugal and tidal acceleration relative to the gravitational acceleration between the two balls, we find that their ratio is proportional to the orbital periods squared
\begin{equation}\label{a_ratio}
\frac{\ddot{r}_{\mathrm{\tau}}}{\ddot{r}_\mathrm{G}} \propto \frac{P^2}{{P_\odot}^2} \approx 10^{-7},
\end{equation}
where $P_\odot$ corresponds to one year, whereas $P$ is the orbital period of the two balls. So tidal and centrifugal effects are about $10^{-7}$ times smaller than the respective gravity of the two balls.

\textbf{Gravitational field:}
Much attention has to be paid to the effect of the gravitational field of the spacecraft itself and the measurement apparatus. This certainly can't be neglected, but with a clever design, its effects can be minimized. For example an almost spherically symmetric spacecraft would probably be a good choice. It should also be possible to calculate the gravitational field caused by the components of the spacecraft at the positions of the balls. Then one can correct the equation of motion by adding an extra term that is accounting for the gravitational field. \\
A source of an additional gravitational field is Earth's moon and the other planets in the solar system. Fortunately, for this we don't need to correct if the whole system, consisting of the spacecraft and the two-body system, is moving through space together. Although, the spacecraft should be able to correct its position around the two balls to avoid a collision.

@@

These are actually three coupled equations of motion, one for each spatial coordinate. Since all are second order differential equations, we need to give six initial conditions to determine the orbit.\\
We choose our coordinate system such that the orbital plane of the two-body system is in the $x-y$ plane. $\vec{\omega}$ is a constant vector that points in the direction of the rotation axis of the Earth-Sun reference frame. We will fit for $\vec{\omega}$, so we don't have to know it exactly before, but it can be estimated well and thus provides good starting points for the fit. The same is valid for the six components of the matrix $C$.
We choose our coordinate system such that the orbital plane of the two-body system is in the $x-y$ plane. $\vec{\omega}$ is a constant vector that points in the direction of the rotation axis of the Earth-Sun reference frame. We will fit for $\vec{\omega}$, so we don't have to know it exactly before, but it can be estimated well and thus provides good starting points for the fit. The same is valid for the six components of the matrix $C$.

@@

In the next step, we set the parameters to a fixed ``true" value
within the range estimated. We have to keep in mind that $k$, $x_0$
and $v_{y_0}$ are linked together and there must not be contradictions
in the true values. The chosen values are displayed in the right
column of Table~\ref{tab:parameters}.  To simulate data we would
obtain from the position measurements, we integrate the equations of
motion. The time for one orbit without perturbations can be estimated
using Kepler's third law (\ref{eq:kepler3}). For the value of $k$
assumed above and a semi-major axis of $0.15$~m, this
results in an orbital period of roughly 12900 seconds or
3.5 hours. Here we integrate two orbits in 100000
steps, that would correspond to about 4 measurements per second,
which should be feasible. To the generated data, we add Gaussian noise
with a standard deviation of 10~nm. The value we use is just an estimate,
it would depend strongly on the measurement system that has been used.

@@

\section{Conclusion}
We have outlined an orbit experiment to measure the gravitational
constant $G$, and estimated the level of various disturbing effects
and how to correct for them.  Simulating the measurement process and
extraction of the gravitational constant indicates that $G$ could be
estimated to about $10^{-7}$ relative uncertainty.  This is about a
factor of 300 better than today's recommended CODATA value. A further increase in precision could be obtained by running the experiment longer. Measuring for 200 instead of two orbits could increase the precision by a factor of 10.
Hence, about the same relative uncertainty as for Planck's constant
$h$ \citep[$1.2 \cdot 10^{-8}$][]{CODATA17} could be achieved. This then
would allow a future redefinition of the kilogram as an integer
multiple of the Planck mass.

>>>>

\abstract
% context heading (optional)
% {} leave it empty if necessary  
{As the kilogram assumes its new definition in terms of $h,c,$ and the
  standard second, it is interesting to think about a possible future
  definition entirely in terms of fundamental constants.  A kilogram
  could, in principle, be defined as an integer multiple of the Planck
  mass (or 4594????$\,\times\sqrt{\hbar c/G}$) without changing it
  currently realizable value.  The practical hurdle is measuring $G$
  four orders of magnitude better than now known, in order to fix the
  four unknown digits.}
% aims heading (mandatory)
{We assess the prospects for measuring $G$ to the desired accuracy
  using an artificial two-body orbit.}
% methods heading (mandatory)
{We consider the simplest configuration, which would be a system
  placed on a Trojan orbit to minimize external gravitational
  perturbations, and enclosed in a cavity chilled to liquid-helium
  temperatures, to suppress radiation pressure.}
% results heading (mandatory)
{Data analysis would involve marginalizing over at least 15~nuisance
  parameters (6~for orbital parameters and 9~for fictitious forces)
  but that, we show, is feasible.}
% conclusions heading (optional), leave it empty if necessary 
{Artificial two-body orbits merit serious consideration as a method
  for measuring $G$.  The configuration considered in this preliminary
  study may be unrealistically expensive, but more complicated but
  more affordable scenarios in Earth orbit may be practical.}

@@

The reform of SI units to use quantum standards is an ongoing ``news
story'' in physics.\footnote{International Bureau of Weights and
  Measures, ``DRAFT 9th edition of the SI Brochure", BIPM.}  At the
heart of the reform are two quantum phenomena that appear on
macroscopic scales: the Josephson effect makes $h/e$ measurable as a
voltage-to-frequency ratio, while the quantum Hall effect makes
$h/e^2$ a standard resistance (the von~Klitzing constant).  These
enable electrical units to be defined in terms of $h$ and $e$.  A
further ingenious instrument, known as a Kibble balance or watt
balance, precisely compares electrical and mechanical power, thus
enabled mass to be standardized in terms of $h$.  For a simple
introduction to the new SI, we refer the reader to the article on the
``LEGO watt balance'' \citep{Chao15}. The proposed redefinition of the
kilogram is as an explicit value for $h$ and $c$, but we can think of
it as
\begin{equation}
\mathrm{kg} \equiv 8.52246550 \times 10^{50} \, \mathrm{s}^{-1}
                   \times c^{-2} \hbar .
\end{equation}
As we can see, the kilogram is coupled to the definition of a second.
That is not a problem in practice, as the second is standardized
orders of magnitude better than any other unit.  Nonetheless, it is
interesting to consider a logical further step of decoupling the
kilogram from other definitions.  This could be done by introducing
the gravitational constant, since $\sqrt{hc/G}$ has dimensions of
mass.  Historically, there have been several proposed unit systems
that included $G$ in some way \citep[for a review,
  see][]{Tomilin1999}, but the idea of including both gravity and
quantum mechanics in a unit system comes from \cite{Planck99}. In the
process of working towards the radiation law, Planck realized that
$c,G,h$ and the Boltzmann constant $k_B$ (in his notation
$G\rightarrow f$, $h\rightarrow b$, $h/k_B\rightarrow a$) naturally
define units for mass, length, time, and temperature.  Indeed, Planck
remarked that such units would be meaningful even in non-human and
extraterrestrial cultures.

A hypothetical Planckian SI unit system would have
\begin{equation} \label{eq:defs}
\begin{aligned}
\mathrm{kg} &= 4.594\ldots \times 10^7
               \left(\frac{\hbar c}{G}\right)^{\!1/2} \\
\mathrm{m}  &= 6.187\ldots \times 10^{34}
               \left(\frac{\hbar G}{c^3}\right)^{\!1/2} \\
\mathrm{s}  &= 1.854\ldots \times 10^{43}
               \left(\frac{\hbar G}{c^5}\right)^{\!1/2} \\
\mathrm{K}  &= 7.058\ldots \times 10^{-33}
               \left(\frac{\hbar c^5}{Gk_B^2}\right)^{\!1/2}
\end{aligned}
\end{equation}
with the numerical coefficients chosen so as to match the present
standards.  The square-rooted expressions are conventionally known as
the Planck mass, Planck length, and so on.  (They are actually
$\sqrt{2\pi}$ smaller than Planck's definitions.)  Extension to
electrical units is straightforward, but we will not discuss it here.
The Planck length, time, and temperature in themselves are extreme
scales, not probed by any known physical process, and presumably the
regime of quantum gravity.  But extreme scales do not harm if they are
only part of some expression, as they are here.  The Planck mass is
small but microscopically measurable: $\sim20\,\mu\mathrm{g}$.  This
is of the same order as the uncertainty in the old prototype
kilogram\footnote{\url{<https://www.bipm.org/utils/common/pdf/cms_m.pdf>}}
and the uncertainty in watt-balance experiments.  Hence the kilogram
could be redefined as an integer multiple of the Planck mass, without
significantly perturbing its value.

@@

\subsection{An artificial two-body gravitating system}
Newton's law of gravitation explains not only why an apple falls down
from a tree, but also orbits. If one has two bodies of masses $m_1$
and $m_2$, on which the only force acting is their respective
gravitational attraction, they will orbit around the common center of
mass, with the relative vector tracing an ellipse described by
Kepler's laws. The third law, relates the orbital period $P$ to
semimajor axis $a$ of the ellipse, as
\begin{equation}\label{eq:kepler3}
P^2 = \frac{4 \, \pi^2}{G(m_1+m_2)}a^3
\end{equation}
with the gravitational constant $G$ appearing in the proportionality
factor.  If one can measure the position of the two masses very
accurately and thus the orbit is known, it is possible to deduce the
gravitational constant.  So what we need is a freely-falling two-body
system. The only way to achieve this over some time, is to go into
space. Thus, one has to design a spacecraft with a cavity that
contains the two masses and a position measurement apparatus. The
cavity should shield the two-body system from environmental
influences. The spacecraft is then put in space, at a location where
tidal forces are minimal.  The stable Lagrange points of the Earth-Sun
system (that is, $L_4$ or $L_5$, also known as the Earth-Trojan
locations) would be suitable.

@@

\textbf{Radiation pressure:} The envelope of the cavity needs to protect the two body-system from the constant bombardment of radiation and charged particles from the Sun. Since the activity of the Sun varies much, the spacecraft must be able to correct its position around the freely falling masses inside the cavity.\\
Furthermore, there is thermal radiation inside the cavity, that could disturb the orbit.
Using the Stefan-Boltzmann law, we can relate radiation pressure to the temperature inside the cavity. If we then specify to balls of diameter $d$, we can express the force on the balls as 
\begin{equation}
F = \frac{2 \pi^6 d^2 {k_\mathrm{B}}^4 T^4}{45 c^3 h^3}.
\end{equation}
As we can see, this force is highly temperature dependent. In order to minimize this influence, one should deploy a temperature as small as possible in the cavity. In fact, if a cooling system based on liquid helium is installed, temperature can be brought down to about 4 K. Such a low temperature results in a force on the balls in the order of $10^{-16}$~N.   This is $\sim10^{-9}$ of the gravitational force, and thus we don't have to further model the effects of radiation pressure.

@@

\textbf{Centrifugal and tidal forces:}
\label{sec:tidal}
The free-fall condition is not possible to achieve for both masses simultaneously since we are in a non-inertial, rotating frame and therefore have to account for centrifugal forces. The centrifugal acceleration $\ddot{\vec{r}}_{\mathrm{Z}}$ is proportional to the distance vector $\vec{r}$ and scales with $\omega^2$, where $\omega$ is the angular velocity of the Sun-Earth reference frame:
\begin{equation}
\label{eq:centrifugalacc}
\ddot{\vec{r}}_{\mathrm{Z}} = \vec{\omega} \times \left( \vec{\omega} \times \vec{r} \right) 
\end{equation}
In addition, there are tidal forces. If one mass is closer to the Sun, the gravitational force on it is slightly bigger than on the other mass. This results in a force perturbation, which is also proportional to $r$. One can approximate the tidal acceleration as
\begin{equation}
\label{eq:tidalacc}
\ddot{\vec{r}}_{\mathrm{\tau}} \approx \frac{2 \, G \, M_\odot}{R^3} \, \vec{r}.
\end{equation}
In this equation $M_\odot$ is the solar mass and $R$ is the radius of the orbit around the Sun. With Kepler's third law one can rewrite the tidal acceleration as a function of the orbital period squared of the Sun-Earth reference frame. Thus it scales the same way as the centrifugal force. To estimate the magnitude of the centrifugal and tidal acceleration relative to the gravitational acceleration between the two balls, we find that their ratio is proportional to the orbital periods squared
\begin{equation}\label{a_ratio}
\frac{\ddot{r}_{\mathrm{\tau}}}{\ddot{r}_\mathrm{G}} \propto \frac{P^2}{{P_\odot}^2} \approx 10^{-7},
\end{equation}
where $P_\odot$ corresponds to one year, whereas $P$ is the orbital period of the two balls. So tidal and centrifugal effects are about $10^{-7}$ times smaller than the respective gravity of the two balls.

\textbf{Gravitational field:}
Much attention has to be paid to the effect of the gravitational field of the spacecraft itself and the measurement apparatus. This certainly can't be neglected, but with a clever design, its effects can be minimized. For example an almost spherically symmetric spacecraft would probably be a good choice. It should also be possible to calculate the gravitational field caused by the components of the spacecraft at the positions of the balls. Then one can correct the equation of motion by adding an extra term accounting for the gravitational field.

A source of an additional gravitational field is the Earth's moon and the other planets in the solar system. Fortunately, for this we don't need to correct if the whole system, consisting of the spacecraft and the two-body system, is moving through space together. However, the spacecraft should be able to correct its position around the two balls to avoid a collision.

@@

These are actually three coupled equations of motion, one for each spatial coordinate. Since all are second order differential equations, we need to give six initial conditions to determine the orbit.

We choose our coordinate system such that the orbital plane of the two-body system is in the $x-y$ plane. $\vec{\omega}$ is a constant vector that points in the direction of the rotation axis of the Earth-Sun reference frame. We will fit for $\vec{\omega}$, so we don't have to know it exactly before, but it can be estimated well and thus provides good starting points for the fit. The same is valid for the six components of the matrix $C$.

@@

In the next step, we set the parameters to a fixed ``true" value
within the range estimated. We have to keep in mind that for given $k$
and $x_0$, $v_{y_0}$ must be chosen to as to produce a nearly-circular
orbit, as desired.  The chosen values are displayed in the right-hand
column of Table~\ref{tab:parameters}.  To simulate data we would
obtain from the position measurements, we integrate the equations of
motion. The time for one orbit without perturbations can be estimated
using Kepler's third law (\ref{eq:kepler3}). For the value of $k$
assumed above and a semi-major axis of $0.15$~m, this results in an
orbital period of roughly 12900 seconds or 3.5 hours. Here we
integrate two orbits in 100000 steps, that would correspond to about 4
measurements per second, which should be feasible. To the generated
data, we add Gaussian noise with a standard deviation of 10~nm. The
value we use is just an estimate, it would depend strongly on the
measurement system that has been used.

@@

\section{Conclusion}
We have outlined an orbit experiment to measure the gravitational
constant $G$, and estimated the level of various disturbing effects
and how to correct for them.  Simulating the measurement process and
extraction of the gravitational constant indicates that $G$ could be
estimated to about $10^{-7}$ relative uncertainty.  This is about a
factor of 300 better than today's recommended CODATA value. A further
increase in precision could be obtained by running the experiment
longer. Measuring for 200 instead of two orbits could increase the
precision by a factor of 10.  Hence, about the same relative
uncertainty as for Planck's constant $h$ \citep[$1.2 \cdot 10^{-8}$,
  see][]{CODATA17} could be achieved. This then would allow a future
redefinition of the kilogram as an integer multiple of the Planck
mass.
